### 2010-03-11 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: Code and interpretation face to face, for all Desc and IDesc definitions.
Desc@ICFP: finish IDesc section.

\begin{structure}
!!! Need Help !!!
<- Motivation
    <- Desc: expressivity of simply-typed data-types: inductive types
        <- Values do not influence types
    /> Example: Vectors
        <- Cannot be defined by just induction
            <- Vectors of all size need to be defined at the *same* time
            -> Defined as a *family* of types
                -> Index
        -> I -> IDesc I: Inductive family
    ???
\end{structure}

@@

\begin{structure}
<- Labelling Id
    <- We had: data Desc : Set -> Set
    -> We want: data IDesc : (I -> Set) -> Set
        <- Indexed functor (?)
        -> It is sufficient to label Id
            <- Where the functor is built
\end{structure}

\begin{structure}
<- Also replacing '1 by 'const  [figure]
    <- For convenience
        <- 'const X equivalent to 'sigma X (\_ -> '1)
        /> Easier to abstract
            <- Extensionally same
            /> 'const more useful in practice
\end{structure}

\[
\stk{
\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
\]

\[\stk{
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To \Set \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}\]

\subsection{Examples}

\begin{structure}
<- Nat
    -> [equation]
    <- Non-indexed types lives in IDesc 1
        -> This applies to all previous examples
\end{structure}

\begin{structure}
<- Levitation [figure]
    <- Following Desc encoding
        /> Note: simple data-type
            -> Live in IDesc 1
    -> Behind the scene, relies on the special purpose switchD
\end{structure}

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
\DConst{},
                                                                                    \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} ]) \\
                              & \left[\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             \right]
\end{array}
}\]

\begin{structure}
<- Untyped lambda terms [figure]
    <- Indexed by Nat
        <- De Bruijn "encoding"
    <- Variable in Fin n
        -> Correct by construction lambda term
    /> Lack the monadic structure [Altenkirch & Reus ???]
        -> Miss substitution operator
        /> Will be fixed later
\end{structure}

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
                            & \left[\begin{array}{l}
                                  \DConst{\Fin{n}} \\
                                  \DProd{\DVar{n}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                              \end{array}\right]
\end{array}
}\]

@@

\begin{structure}
!!! Need Help !!!
<- Assuming a suitable notion of definitional equality for I
    ???
\end{structure}

\begin{structure}
<- Fin [figure]
    <- Presentation with equality constraints
    <- Example of constrained constructor
        -> Constraints are translated into equalities
    ???
\end{structure}

@@

\begin{structure}
<- Vectors
    Do we treat them in the end? 
    What can we say here we haven't with Fin?
\end{structure}

\begin{structure}
<- GADT-style data-type definition
    <- We directly support it
        <- Constraints turned into equalities
            <- Fin example
    <- Again, smooth transition from "today" data-types to advanced ones
        <- Same grammar
        /> More power
    /> Equations in data-types are not a blessing
        <- Complexify induction principle
        <- Lose match between data-type definition and the elimination form
\end{structure}

\begin{structure}
!!! Need Help !!!
<- Brady optimization: forcing
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
    ??? More technical detail needed
\end{structure}

@@

@@

\begin{structure}
<- Variation on a theme: free imonad construction
    <- Recall existence of generic free monad construction
    -> Present its generalization to IDesc [equation]
        <- \I -> IDesc I as describing an indexed endofunctor
        <- Free indexed monad construction
    -> Still a suitable, generic notion of substitution
        <- show type signature
        <- show implementation?? (space! space!)
\end{structure}

\begin{structure}
<- Untyped lambda terms
    <- Obvious instance of free monad
        <- Variable: Return
        <- App and Abst: indexed functor
        <- Bind: substitution
    -> Get a correct-by-construction lambda term
        /> With substitution for free!
        <- We should not pay for the structure which *is* here
\end{structure}

@@

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\LambdaTD\: \mapsto \FreeMonad{[\LambdaTFreeD]}{\Fin{}}
}\]

\begin{structure}
<- IDesc
    <- Another instance of free monad
        <- Var: Return
        <- Remaining: (not even indexed) functor
    -> Madness just starts
        <- map operation
        <- [ subst sigma D ] X = [D] (\x -> [subst sigma x] X)
        ???
\end{structure}

@@

@@

>>>>

\begin{wstructure}
!!! Need Help !!!
<- Motivation
    <- Desc: expressivity of simply-typed data-types: inductive types
        <- Values do not influence types
    /> Example: Vectors
        <- Cannot be defined by just induction
            <- Vectors of all size need to be defined at the *same* time
            -> Defined as a *family* of types
                -> Index
        -> I -> IDesc I: Inductive family
    ???
\end{wstructure}

So far, we have explored the well-known realm of inductive types. We
have built upon the experience gained in using simply-typed languages,
such as Haskell or OCaml. In our dependent setting, we provided these
data-types by the mean of $\Desc$, a universe of descriptions.

While inductive types are the alpha and omega of the simply-typed
languages, evolving in a dependent setting fosters new
opportunities. The typical example is bounded lists, also known as
vectors. A vector is a list decorated by its length. Having this
information prevents hazardous operations, such as taking the head of
an empty vector: the head function only takes vectors of length
$\NatSuc{n}$, as enforced by its type. This is made possible by the
very specificity of dependently-typed systems: terms -- the length --
are allowed to influence types -- the vector type.

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, we adopt inductive family of types as our basic
grammar of data-types. To this end, we rely on \emph{indexing}. We are
going to transform the $\Desc$ universe into an indexed $\IDesc$
universe. We express inductive families by the $I \To \IDesc{I}$ type.

@@

\begin{wstructure}
<- Labelling Id
    <- We had: data Desc : Set -> Set
    -> We want: data IDesc : (I -> Set) -> Set
        <- Indexed functor (?)
        -> It is sufficient to label Id
            <- Where the functor is built
\end{wstructure}

In the previous section, we have presented the $\Desc$ universe as the
grammar of functors in the category $\Set$. We have seen how to encode
inductive types in this setting. To encode an inductive family indexed
by $\Bhab{\V{I}}{\Set}$, we rely on functors in the category
$\Set^I$. We call these \emph{indexed functors}. When $\Desc$ was
defining a grammar a functor, we need $I \To \IDesc{I}$ to define a
grammar of indexed functors:

\[\stk{
\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

This generalization can be achieved by a minor evolution of $\Desc$
(Fig.~\ref{fig:factor-desc}). Indeed, the functorial nature of $\Desc$
is entirely captured by the $\DId$ code. We get an indexed functor
by turning $\DId$ into an index-dependent $\DVar$ code:

\[\stk{
\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \ldots          & \ldots
\end{array} \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set        \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\ldots                        &  \ldots
\end{array}
}\]

\begin{wstructure}
<- Also replacing '1 by 'const  [figure]
    <- For convenience
        <- 'const X equivalent to 'sigma X (\_ -> '1)
        /> Easier to abstract
            <- Extensionally same
            /> 'const more useful in practice
\end{wstructure}

The resulting code of $\IDesc$ is presented in
Figure~\ref{fig:idesc}. The reader will notice that we have also
replaced $\DUnit$ by a more general $\DConst$ code. Whereas $\DUnit$
was interpreted as the unit set, $\DConst{X}$ is interpreted as $X$,
for any $\Bhab{X}{\Set}$. Extensionally, $\DConst{X}$ and
$\DSigma{X}{\DUnit}$ are equivalent. However, $\DConst$ is more
succinct. More importantly, $\DConst$ is \emph{first-order}, unlike
its equivalent encoding. From a definitional perspective, we are
giving more opportunities to the type-system, hence reducing the
burden on the programmer in practice.

\begin{figure*}

\[
\begin{array}{ll}
\stk{
\data \IDesc{} : \PI{\V{I}}{\Set} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:}l@{\quad}l}
    \DVar{}         & I \To \IDesc{I}                                   \\
    \DConst{}       & \Set \To \IDesc{I}                                \\
    \DProd{}{}      & \PI{\V{D}, \V{D'}}{\IDesc{I}} \IDesc{I}           \\
    \DSigma         & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I}  \\
    \DPi            & \PI{\V{S}}{\Set} \PIS{S \To \IDesc{I}} \IDesc{I} 
\end{array}
}
&
\stk{
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{I} \To (\V{I} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{i}}{I}{P}      &  P~i                                                 \\
\idescop{\DConst{X}}{I}{P}    &  X                                                   \\
\idescop{\DProd{D}{D'}}{I}{P} &  \TIMES{\idescop{D}{I}{P}}{\idescop{D'}{I}{P}}       \\
\idescop{\DSigma{S}{D}}{I}{P} &  \SIGMA{\V{s}}{S} \idescop{D\: s}{I}{P}                    \\
\idescop{\DPi{S}{D}}{I}{P}    &  \PI{\V{s}}{S} \idescop{D\: s}{I}{P}            
\end{array}
}
\end{array}
\]

\caption{Universe of indexed description}
\label{fig:idesc}

\end{figure*}

\subsection{Examples}
\label{sec:idesc-examples}

\begin{wstructure}
<- Nat
    -> [equation]
    <- Non-indexed types lives in IDesc 1
        -> This applies to all previous examples
\end{wstructure}

In order to gain some intuition with $\IDesc$, let us re-implement the
signature functor of natural numbers. We refer the reader to
Section~\ref{sec:desc-examples} for its incarnation in the universe of
descriptions. The code is the following:

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT [ \NatZero, \NatSuc{} ])}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

Because $\Nat$ is just an inductive type, we make no use of the index:
it is the trivial inhabitant of $\Unit$. Therefore, the recursive
argument is materialized by $\DVar{\Void}$, where we were using $\DId$
in the $\Desc$ presentation. This transformation applies to all
inductive types: we have not lost in expressive power during the
transition.

\begin{wstructure}
<- Levitation [figure]
    <- Following Desc encoding
        /> Note: simple data-type
            -> Live in IDesc 1
    -> Behind the scene, relies on the special purpose switchD
\end{wstructure}

A mandatory exercise consists in describing $\IDesc$ in itself. With
the experience gained in Section~\ref{sec:desc-levitate}, this
levitation does not pose any issue:

\[\stk{
\IDescD : \PI{\V{I}}{\Set} \IDesc{\Unit} \\
\begin{array}{@{}ll}
\IDescD~I \mapsto \DSigma{}{} & (\EnumT [ \DVar{},
                                          \DConst{},
                                          \DProd{}{},
                                          \DSigma{}{}, 
                                          \DPi{}{} ]) \\
                              & \left[\begin{array}{l}
                                      \DConst{I}                  \\
                                      \DConst{\Set}               \\
                                      \DProd{\DVar{\Void}}{\DVar{\Void}}  \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})} \\
                                      \DSigma{\Set}{(\LAM{S} \DPi{S}{(\LAM{\_} \DVar{\Void})})}
                                   \end{array}
                             \right]
\end{array}
}\]

Note that this description also lives in $\IDesc{\Unit}$. Indeed,
$\IDesc$ is merely an inductive type. As for the $\Desc$ levitation,
the finite function space $[ \ldots ]$ elaborates into a special
purpose $\F{switchID}$ operator.

\begin{wstructure}
<- Untyped lambda terms [figure]
    <- Indexed by Nat
        <- De Bruijn "encoding"
    <- Variable in Fin n
        -> Correct by construction lambda term
    /> Lack the monadic structure [Altenkirch & Reus ???]
        -> Miss substitution operator
        /> Will be fixed later
\end{wstructure}

Our next example is, for once, building an inductive family. We
present the implementation of untyped lambda terms, as a family
indexed by natural numbers. The natural numbers play the role of De
Bruijn indexes. The indexed description is the following:

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\LambdaTD\: n = \DSigma{}{} & (\EnumT [ \LambdaTVar, \LambdaTApp, \LambdaTLam ]) \\
                            & \left[\begin{array}{l}
                                  \DConst{\Fin{n}} \\
                                  \DProd{\DVar{\NatSuc{n}}}{\DVar{n}} \\
                                  \DVar{(\NatSuc{n})}
                              \end{array}\right]
\end{array}
}\]

A variable is any number between $0$ and $n$, hence forbidding the
reference of out-of-scope binders. Making a lambda increases by one
the possible index choice of the underlying term. A function
application applies a term with $n+1$ binders, resulting in a term
with $n$ binders.

We have therefore given a representation of lambda terms as a
data-type. Any inhabitant of this data-type is a valid lambda term,
forbidding out-of-scope reference in variables. While we have the
data-type, we can regret the absence of structure. As shown by
Altenkirch and Reus~\cite{altenkirch:monadic-lambda}, the lambda terms
actually enjoy a monadic structure. In this setting, the ``bind''
corresponds to substitution. We shall remedy to this situation
shortly.

@@

\begin{wstructure}
!!! Need Help !!!
<- Assuming a suitable notion of definitional equality for I
    ???
\end{wstructure}

However, before entering the realm of generic structures, let us
further explore the universe of indexed descriptions. In the
following, we ask $I$ to come equipped with a notion of definitional
equality. Hence, we can legibly equates elements of $I$.

\begin{wstructure}
<- Fin [figure]
    <- Presentation with equality constraints
    <- Example of constrained constructor
        -> Constraints are translated into equalities
    ???
\end{wstructure}

In our implementation of lambda terms, variables were elements of
$\Fin{n}$, with $n$ the number of binders. Let us now implement
$\Fin$. $\Fin$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc$ are only defined for an index distinct from $\NatZero$:

\[
\Rule{\Bhab{n}{\Nat}}
     {\Bhab{\FinZero}{\Fin{(\NatSuc{n})}}}
\qquad
\Rule{\Bhab{n}{\Nat} \quad 
      \Bhab{i}{\Fin{n}}}
     {\Bhab{\FinSuc{}}{\Fin{(\NatSuc{n})}}}
\]

In order to code this data-type, we have to appeal to equality:

@@

Intuitively, the constraint is captured as follow. First, we store an
element $m$ of $\Nat$, thanks to a $\DSigma$. However, the constraint
stipulates that $m$ cannot be \emph{any} natural numbers: it must be
``the index minus one''. This translates into the constraint $n
\PropEq \NatSuc{m}$.

\begin{wstructure}
<- Vectors
    Do we treat them in the end? 
    What can we say here we haven't with Fin?
\end{wstructure}

%%! I suspect we don't, space-wise

\begin{wstructure}
<- GADT-style data-type definition
    <- We directly support it
        <- Constraints turned into equalities
            <- Fin example
    <- Again, smooth transition from "today" data-types to advanced ones
        <- Same grammar
        /> More power
    /> Equations in data-types are not a blessing
        <- Complexify induction principle
        <- Lose match between data-type definition and the elimination form
\end{wstructure}

The $\Fin$ example is also a well-known example of Generalized
Abstract Data-Type (GADT), in Haskell. It comes as no surprise that
our universe of indexed description supports this restricted class of
dependent data-types. As we have shown with $\Fin$, constrained
constructors are straightforwardly translated into equality
constraints.

While we support the same grammar, indexed descriptions embraces a
much wider class of data-types. Indeed, we pose no constraint apart
from strict positivity. This suggests that, based on the familiar
grammar of GADTs, we can simply lift off the restrictions and
translates the user's input to a code in our universe of indexed
description.

However, one could, quite rightfully, be worried by the introduction
of equations in the picture. In the context of programming, these
equations will come back and haunt us, begging for proofs. Hence,
whereas the user typed a constructor-constrained data-type, the
elimination form will involve equations and their proofs. This
introduces a loophole in our abstraction we should try to address.

\begin{wstructure}
!!! Need Help !!!
<- Brady optimization: forcing
    <- Source to source translation
    <- Able to remove some constraints
    -> Example: Fin [figure]
    ??? More technical detail needed
\end{wstructure}

To get a clue of a solution, let us look back at $\Fin$. We note that
the equation is introduced because we are \emph{storing} the index of
the inductive family, through $m$. Hence the constraint. However,
\emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}. Applying the
\emph{forcing} optimization to our definition of $\Fin$ gives the
following, equivalent definition:

@@

Equations have simply disappeared! We should precise that forcing a
description is not garanteed to remove all constraints. It is subject
of future work to see if constraints can be entirely eradicated, or
presented more conveniently to the user.

Finally, it is worth mentioning that the forcing optimization is a
\emph{source-to-source} transformation of the description.

%%! Actual implementation as a generic function??!

@@

\begin{wstructure}
<- Variation on a theme: free imonad construction
    <- Recall existence of generic free monad construction
    -> Present its generalization to IDesc [equation]
        <- \I -> IDesc I as describing an indexed endofunctor
        <- Free monad construction
    -> Still a suitable, generic notion of substitution
        <- show type signature
        <- show implementation?? (space! space!)
\end{wstructure}

In this section, we develop some more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
a free monad operation on simple descriptions. Building on this
experience, we are going to present its equivalent in the indexed
world. The processus is the same. Namely, given an indexed functor, we
derive the indexed functor coding its free monad:

\[\stk{
\begin{array}{ll}
\FreeIMonad{\_}{} : & \PITEL{\V{I}}{\Set}
                      \PITEL{\V{X}}{\V{I} \To \Set} \\
                    & \PITEL{\V{R}}{\V{I} \To \TagIDesc{I}} \To
                      (\V{I} \To \TagIDesc{I})
\end{array} \\
\FreeIMonad{R}{I}{X} \mapsto
    \LAM{i}
    \pair{\ListCons{\DVar{}}{(\fst{(R\: i)})}}
         {\pair{(\DConst{X\: i})}
               {(\snd{(R\: i)})}{}}{}
}\]

Just as in the universe of descriptions, this construction comes with
an obvious ``return'' and a substitution operation, the ``bind''. Its
type signature is the following:

\newcommand{\substI}{\F{substI}}

\[
\substI : \PITEL{\V{I}, \V{J}}{\Set}
          \PITEL{\sigma}{\V{I} \To \IDesc{\V{J}}}
          \PITEL{\V{D}}{\IDesc{I}} \To
          \IDesc{J}
\]

\begin{wstructure}
<- Untyped lambda terms
    <- Obvious instance of free monad
        <- Variable: Return
        <- App and Abst: indexed functor
        <- Bind: substitution
    -> Get a correct-by-construction lambda term
        /> With substitution for free!
        <- We should not pay for the structure which *is* here
\end{wstructure}

In Section~\ref{sec:idesc-examples}, we were missing the monadic
structure of lambda terms. In particular, we wish we were given a
substitution principle for free. Above, we have generically built such
substitution operation for free monads. Can we be rewarded for our
effort? Let us define the following $\LambdaTFreeD$ description,
largely inspired by $\LambdaTD$:

@@

In essence, we have forgotten variables, just keeping the nodes of
lambda terms: abstraction and application. Our hope is that taking the
free monad of this indexed functor gives us our lambda terms back:

%%! FreeMonad takes a damned tagged description

\[\stk{
\LambdaTD : \Nat \To \IDesc{\Nat} \\
\LambdaTD\: \mapsto \FreeIMonad{[\LambdaTFreeD]}{\Nat}{\Fin{}}
}\]

By a straightforward unfolding of definitions, the reader will
convince herself that we recover the initial definition. However, this
time, we have a substitution operation. In summary, we have done less
work in defining $\LambdaTFreeD$ but we have pointed out its
structure, so, rightfully, we have been rewarded in our effort. The
lesson is that, when structure there is, one should be allowed to get
it for free.

\begin{wstructure}
<- IDesc
    <- Another instance of free monad
        <- Var: Return
        <- Remaining: (not even indexed) functor
    -> Madness just starts
        <- map operation
        <- [ subst sigma D ] X = [D] (\x -> [subst sigma x] X)
        ???
\end{wstructure}

Another instance of free monad is $\IDesc$ itself. Indeed, $\DVar$ is
nothing but the ``return''. The remaining constructors are simply the
freed functor, trivially indexed by $\Unit$. Hence, amusingly, we can
define the $\IDesc$ freed functor by the following code:

@@

Then, we get $\IDesc$ by building its free monad:

@@

%%! Should we stop here? Or show off some madness?
